Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
initial abundance

In the original formulation, initial abundance was
modeled using either the Poisson or negative binomial
distributions. Hence, in the Poisson case, we have
\begin{equation}
  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}
where $\Lambda_i$ is the expected abundance at site $i$ during year 1.
From hereout, we will reserve bracket notation for arbitrary
probability distributions, and for simplicity, use the equivalent notation $N_{i1}
\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
%\begin{gather}
%N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
%\text{or} \nonumber \\
%N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
%\label{eq:N1}
%\end{gather}
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution, $N_{i1} \sim
\text{NB}(\Lambda, \alpha)$, allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible, the general expression of which can be written:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
  \label{eq:like}
\end{equation}
This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.

>>>>

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
%\begin{equation}
%  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
where $\Lambda_i$ is the expected abundance at site $i$ during year
1. In the Poisson case, the eqivalent expression of
Eq.~\ref{eq:ss2a} is $[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%From hereout, we will reserve bracket notation for arbitrary
%probability distributions, and for simplicity, use the equivalent notation $N_{i1}
%\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution, $N_{i1} \sim
\text{NB}(\Lambda, \alpha)$, allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.
Note that $N_{i1} \sim \text{Pois}(\Lambda)$ is equivalent to
$[N_{i1} | ] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible, the general expression of which can be written:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i1}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [n_{it}|N_{it}][N_{it}|N_{it-1}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{equation}
This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.
